---
title: "Average Treatment Effect Estimation"
author: Klaus Kähler Holst
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    html-math-method: mathjax
vignette: >
  %\VignetteIndexEntry{Average Treatment Effect Estimation}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  echo: true
  warning: false
  collapse: true
  comment: "#>"
---

```{r, include = FALSE}
library("targeted")
```

# Introduction

Consider observed data $(Y, A, W)$ where $Y$ is the outcome, $A \in \{0,1\}$
is the binary treatment, and $W$ is a vector of covariates. Under the potential
outcomes framework, we define the *average treatment effect* (ATE) as

$$\tau = E[Y(1)] - E[Y(0)]$$

where $Y(a)$ denotes the outcome that would have been observed under treatment
$A = a$. Identification of $\tau$ from observed data requires standard causal
assumptions: consistency, positivity, and no unmeasured confounding.

The `cate` function in the `targeted` package estimates the ATE using the
*augmented inverse probability weighted* (AIPW) estimator, also known as the
doubly-robust estimator. For each treatment level $a$, the efficient influence
function for the marginal mean $\psi_a = E[Y(a)]$ is

$$\varphi_a(Y, A, W) = \frac{I(A=a)}{\pi_a(W)}\{Y - \mu_a(W)\} + \mu_a(W) - \psi_a$$

where $\pi_a(W) = P(A = a \mid W)$ is the propensity score and $\mu_a(W) = E[Y
\mid A = a, W]$ is the outcome regression. The resulting estimator is
*doubly robust*: it is consistent if either $\pi_a$ or $\mu_a$ is correctly
specified. When both models are correctly specified, the estimator achieves the
semiparametric efficiency bound.

# Continuous outcome

## Simulated data

We generate data from a model with a known treatment effect. The outcome
depends on confounders $W_1$ and $W_2$, with an interaction between treatment
and $W_2$:

$$
Y = \cos(W_1) + W_2 A + 0.2 W_2^2 + A + \varepsilon, \quad \varepsilon \sim N(0,1)
$$

The true ATE is $E[W_2 + 1] = 1$ since $W_2 \sim N(0,1)$.

```{r sim-continuous}
sim_data <- function(n = 2000) {
  w1 <- rnorm(n)
  w2 <- rnorm(n)
  a <- rbinom(n, 1, plogis(-1 + w1))
  y <- cos(w1) + w2 * a + 0.2 * w2^2 + a + rnorm(n)
  data.frame(y = y, a = a, w1 = w1, w2 = w2)
}

set.seed(2025)
d <- sim_data(2000)
```

## Parametric estimation

To estimate the ATE, we call `cate` with `cate.model = ~1` (an intercept-only
model, which targets the marginal ATE rather than conditional effects). We
specify parametric models for both the outcome regression and the propensity
score:

```{r ate-parametric}
fit <- cate(
  response.model = y ~ a * (w1 + w2),
  treatment.model = a ~ w1 + w2,
  cate.model = ~1,
  data = d
)
fit
```

The output shows estimates of the potential outcome means $E[Y(1)]$ and
$E[Y(0)]$, followed by the ATE (labelled `(Intercept)`). With correctly
specified parametric models and no cross-fitting, the AIPW estimator is
$\sqrt{n}$-consistent and asymptotically normal.

By default, `cate` includes a second-order correction term (`second.order =
TRUE`) that improves robustness to misspecification of the outcome model when
the propensity model is a GLM.

## Cross-fitting

When flexible or data-adaptive models are used for the nuisance parameters, the
AIPW estimator can suffer from overfitting bias. *Cross-fitting* (sample
splitting) resolves this: the data are partitioned into $K$ folds, nuisance
models are fitted on $K-1$ folds and predictions are made on the held-out fold.
This relaxes the Donsker conditions that would otherwise be required for
$\sqrt{n}$-consistency.

With parametric models, cross-fitting is not strictly necessary but introduces
no harm:

```{r ate-crossfit}
fit_cf <- cate(
  response.model = y ~ a * (w1 + w2),
  treatment.model = a ~ w1 + w2,
  cate.model = ~1,
  data = d,
  nfolds = 5
)
fit_cf
```

## Flexible nuisance models

One of the main strengths of the AIPW framework is that nuisance models can be
replaced with flexible or machine learning estimators without sacrificing valid
inference for the target parameter. The `learner` class in `targeted` provides a
unified interface for specifying these models.

Here we use a generalized additive model (GAM) for the outcome regression,
stratified by treatment arm, combined with a logistic GLM for the propensity
score. Stratification (`stratify = TRUE`) fits separate outcome models for
treated and untreated units:

```{r ate-gam}
fit_gam <- cate(
  response.model = learner_gam(y ~ s(w1) + s(w2)),
  treatment.model = learner_glm(a ~ w1 + w2, family = binomial),
  cate.model = ~1,
  data = d,
  nfolds = 5,
  stratify = TRUE
)
fit_gam
```

For even greater flexibility, an ensemble learner (superlearner) can combine
multiple candidate models. The superlearner selects the optimal convex
combination of learners via cross-validation:

```{r ate-sl}
outcome_model <- learner_sl(
  list(
    glm = learner_glm(y ~ w1 * w2),
    gam = learner_gam(y ~ s(w1) + s(w2))
  ),
  nfolds = 5
)

fit_sl <- cate(
  response.model = outcome_model,
  treatment.model = learner_glm(a ~ w1 + w2, family = binomial),
  cate.model = ~1,
  data = d,
  nfolds = 5,
  stratify = TRUE
)
fit_sl
```

# Binary outcome

The AIPW estimator applies equally to binary outcomes. Here the ATE is a *risk
difference*: the difference in marginal probabilities of the event under
treatment versus control.

## Simulated data

We generate binary outcomes from a logistic model with a true log-odds ratio of
$0.8$:

```{r sim-binary}
sim_binary <- function(n = 2000) {
  w1 <- rnorm(n)
  w2 <- rnorm(n)
  a <- rbinom(n, 1, plogis(-0.5 + 0.5 * w1))
  p1 <- plogis(-1 + 0.5 * w1 - 0.3 * w2 + 0.8 * a)
  y <- rbinom(n, 1, p1)
  data.frame(y = y, a = a, w1 = w1, w2 = w2)
}

set.seed(2025)
db <- sim_binary(3000)
```

## Estimation

```{r ate-binary}
fit_bin <- cate(
  response.model = learner_glm(y ~ a * (w1 + w2), family = binomial),
  treatment.model = learner_glm(a ~ w1 + w2, family = binomial),
  cate.model = ~1,
  data = db,
  nfolds = 5
)
fit_bin
```

The output shows the estimated marginal event probabilities $E[Y(1)]$ and
$E[Y(0)]$ and their difference (the risk difference). To obtain an odds ratio,
we can transform the potential outcome estimates using `lava::estimate`:

```{r ate-or}
logit <- lava::logit
or <- lava::estimate(fit_bin$estimate, function(p) logit(p[1]) - logit(p[2]),
                     labels = "log(OR)")
merge(or, exp(or), labels = c("log(OR)", "OR"))
```

# Conditional average treatment effects

The examples above all use `cate.model = ~1`, targeting the marginal ATE. The
`cate` function also supports estimation of *conditional* average treatment
effects (CATE) by specifying a richer projection model. For instance,
`cate.model = ~1 + w2` projects the individual-level treatment effect onto a
linear model in $W_2$, allowing the estimated effect to vary across levels of
$W_2$. See `?cate` for details and examples.

# Session info

```{r sessioninfo}
sessionInfo()
```
