---
title: Prediction model class (`learner`)
author: Klaus Kähler Holst, Benedikt Sommer
date: "`r Sys.Date()`"
output:
  knitr:::html_vignette:
    fig_caption: yes
    fig_width: 5.15
    fig_height: 3.5
    fig_align: center
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Prediction models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
 collapse = TRUE,
 #dev="png",
 comment = "#>"
)
library("targeted")
```

# Introduction

Methods for targeted and semiparametric inference rely on fitting nuisance
models to observed data when estimating the target parameter of interest. The
`{targeted}` package implements the [R6 reference class](https://r6.r-lib.org/)
`learner` to harmonize common statistical and machine learning models for the
usage as nuisance models across the various implemented estimators. Commonly
used models are constructed as `learner` class objects through the `learner_`
functions. These functions are wrappers for statistical and machine learning
models that have been implemented in other R packages. This is conceptually
similar to packages such as `{caret}` and `{mlr3}`. Besides implementing a large
number of prediction models, these packages provide extensive functionalities
for data preprocessing, model selection and diagnostics, and hyper-parameter
optimization. The design of the `learner` class and accompanied functions is
much leaner, as our use-cases often only require a standardized interface for
estimating models and generating predictions, instead of the full predictive
modelling pipeline.

This vignette uses the Mayo Clinic Primary Biliary Cholangitis data
(`?survival::pbc`) to exemplify the usage of the `learner` class objects. Our
interest lies in the prediction of the composite event of death or transplant
before 2 years.
```{r data}
data(pbc, package="survival")
pbc <- transform(pbc, y = (time < 730) * (status > 0))
```

A logistic regression model with a single `age` covariate is defined and
estimated via

```{r logistic1}
lr <- learner_glm(y ~ age, family = binomial())
lr$estimate(pbc)
```

and predictions for the event `y = 1` (class 2 probabilities) for new data are
obtained by

```{r logistic2}
lr$predict(newdata = data.frame(age = c(20, 40, 60, 80)))
```

The remainder of this vignette is structured as follows. We first provide more
details about the built-in learners that wrap statistical and machine
learning models from other R packages. Once the usage of the returned
`learner` class objects has been introduced, we provide details to implement
new learners.

# Built-in learners

The various constructors for commonly used statistical and machine learning
models are listed as part of the `learner` class documentation
```{r, eval = FALSE}
?learner # help(learner)
```
which contains essential information for the usage of the `learner` class
objects. With the exception of `learner_sl`, all constructors require a
`formula` argument that specify the response vector and design matrix of a
learner. This way the construction of a learner is separated from the estimation
of the underlying model, as shown in the above logistic regression example. A
result of this design is the convenient specification of ensemble learners
(superlearner), where individual learners operate on a different subset of
features.

```{r}
lr_sl <- learner_sl(
  learners = list(
    glm1 = learner_glm(y ~ age * bili, family = "binomial"),
    glm2 = learner_glm(y ~ age, family = "binomial"),
    gam = learner_gam(y ~ s(age) + s(bili), family = "binomial")
  )
)
lr_sl$estimate(pbc, nfolds = 10)
lr_sl
```

## Details
All constructors are essentially wrappers for the fit and predict functions of
the underlying statistical or ML models, with minor modifications to standardize
the output of the predict functions. For example, the fit function
`grf::probability_forest` (generalized random forest) that is wrapped inside
`learner_grf` requires a design matrix `X` and response vector `Y` as inputs.

A key feature of the `learner` class is the internal construction of the design
matrix and response vector from the specified `formula` argument.
```{r}
lr_grf <- learner_grf(y ~ age + log2(bili), model = "probability_forest")
lr_grf$summary()$estimate
```
The during the fitting process constructed design matrix can be inspected via
```{r}
lr_grf$design(pbc)$x |> head()
```

# Usage

cross validation

updating

specials

cloning

# Defining new learners
