---
title: "Prediction model class (`ml_model`)
author: "Klaus Kähler Holst, Benedikt Sommer"
date: "`r Sys.Date()`"
output:
  knitr:::html_vignette:
    fig_caption: yes
    fig_width: 5.15
    fig_height: 3.5
    fig_align: center
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Prediction models - scoring and calibration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
 collapse = TRUE,
 #dev="png",
 comment = "#>"
)
library("targeted")
```

# Introduction

Methods for targeted and semiparametric inference rely on fitting nuisance
models to observed data when estimating the estimand of interest. The
`{targeted}` implements the [R6 reference class](https://r6.r-lib.org/)
`learner` to harmonize common statistical and machine learning models for the
usage as nuisance models across the various implemented estimators. Commonly
used models are instantiated as `learner` class objects through the `learner_`
functions. These functions are wrappers for statistical and machine learning
models that have been implemented in other R packages. This is conceptually
similar to packages such as `{caret}` and `{mlr3}`. Besides implementing a large
number of prediction models, they provide extensive functionalities for data
preprocessing, model selection and diagnostics, and hyper-parameter
optimization. The design of the `learner` class and accompanied functions is
much leaner, as our use-cases often only require a standardized interface for
estimating models and generating predictions, instead of the full predictive
modelling pipeline.

We use the Mayo Clinic Primary Biliary Cholangitis data (`?survival::pbc`)
to exemplify the usage of the `learner` class objects. Our interest is to
predict the composite event of death or transplant before 2 years.
```{r data}
data(pbc, package="survival")
pbc <- transform(pbc, y = (time < 730) * (status > 0))
```

A logistic regression model with a single `age` covariate is defined and
estimated via

```{r logistic1}
lr <- predictor_glm(y ~ age, family = binomial())
lr$estimate(pbc)
```

and predictions of class probabilities for new data are obtained by

```{r logistic2}
lr$predict(newdata = data.frame(age = c(20, 40, 60, 80)))
```

The remainder of this vignette is structured as follows. We first provide more
details about the built-in learners that wrap statistical and machine
learning models from other R packages. Once the usage of the returned
`learner` class objects has been introduced, we provide details to implement
new learners.

# Built-in learners

- start with listing all estimators
- passing arguments to fit and predict function, related to inherited roxygen
documentation (exposing most commonly used arguments)
- 

# Usage

cross validation

updating

specials

cloning

# Defining new learners
