<!-- README.md is generated from README.Rmd. Please edit that file -->


```{r  include=FALSE }
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "50%"
)
```

<!-- badges: start -->
[![r-cmd-check](https://github.com/kkholst/targeted/actions/workflows/r_check.yaml/badge.svg?branch=main)](https://github.com/kkholst/targeted/actions/workflows/r_check.yaml)
[![codecov](https://codecov.io/github/kkholst/targeted/graph/badge.svg?token=MMSEDLWIZd)](https://codecov.io/github/kkholst/targeted)
[![license](https://img.shields.io/badge/License-Apache%202.0-blue.svg)](
  https://opensource.org/license/apache-2-0)
[![cran](https://www.r-pkg.org/badges/version-last-release/targeted)](https://cranlogs.r-pkg.org/downloads/total/last-month/targeted)
<!-- badges: end -->


# Targeted Learning in R: `targeted` <a href="https://kkholst.github.io/targeted"><img src="man/figures/logo.svg" align="right" height="75" /></a>

Various methods for targeted learning and semiparametric inference including
augmented inverse probability weighted (AIPW) estimators for missing data and
causal inference (Bang and Robins (2005)
<doi:10.1111/j.1541-0420.2005.00377.x>), variable importance and conditional
average treatment effects (CATE) (van der Laan (2006)
<doi:10.2202/1557-4679.1008>), estimators for risk differences and relative
risks (Richardson et al. (2017) <doi:10.1080/01621459.2016.1192546>), assumption
lean inference for generalized linear model parameters (Vansteelandt et al.
(2022) <doi:10.1111/rssb.12504>).


# Installation

You can install the released version of targeted from [CRAN](<https://CRAN.R-project.org>) with:

```{r  install, eval=FALSE }
install.packages("targeted")
```

And the development version from [GitHub](<https://github.com/kkholst/targeted>) with:

```{r  eval=FALSE }
remotes::install_github("kkholst/targeted", ref="dev")
```

# Examples

To illustrate some of the functionality of the `targeted` package we simulate
some data from the following model
$$Y = \cos(W_1) + A + W_2 A + 0.2 W_2^2 + \epsilon$$
with independent measurement error $\epsilon\sim\mathcal{N}(0,1)$, and
with treatment variable $A \sim Bernoulli(\operatorname{expit}\{-1+W_1\})$ and
independent covariates $W_1, W_2\sim\mathcal{N}(0,1)$.
```{r simdata}
simdata <- function(n, ...) {
   w1 <- rnorm(n) # covariates
   w2 <- rnorm(n)
   a <- rbinom(n, 1, plogis(-1 + w1)) # treatment
   y <- cos(w1) + w2*a + 0.2*w2^2 + a + rnorm(n) # continuous response
   data.frame(y, a, w1, w2)
}
d <- simdata(1000)

head(d)
```

## Prediction models

Methods for targeted and semiparametric inference rely on fitting nuisance
models to observed data when estimating the target parameter of interest. The
`{targeted}` package implements the [R6 reference class](https://r6.r-lib.org/)
`learner` to harmonize common statistical and machine learning models for the
usage as nuisance models across the various implemented estimators, such as the
`targeted:cate` function. Commonly used models are constructed as `learner`
class objects through the `learner_` functions.



##


## Risk regression with binary exposure and nuisance model for the odds-product regression
Estimating a relative risk or risk difference with a binary exposure


```{r  results='hide' }
library(targeted)
```

Simulate some data:

```{r   }
m <- lvm() |>
    regression(a ~ x+z) |>
    regression(lp.target ~ 1) |>
    regression(lp.nuisance ~ x + z) |>
    distribution('a', binomial.lvm("logit")) |>
    binomial.rr('y', 'a', 'lp.target', 'lp.nuisance')

par <- c('a'=-2, 'lp.target'=1, 'lp.nuisance'=-1, 'lp.nuisance~x'=2)
d <- lava::sim(m, n=1e4, seed=1, p=par) |>
    subset(select=c('y', 'a','x','z'))

head(d)
```

```{r   }
fit <- riskreg(y ~ a, nuisance=~x+z, data=d, type="rr")
fit
```

Here the same design matrix is used for both the propensity model and
the nuisance parameter (odds-product) model

```{r   }
summary(fit)
```

Double-robustness illustrated by using a wrong propensity
model but a correct nuisance paramter (odds-product) model:

```{r   }
riskreg(y ~ a, nuisance=~x+z, propensity=~z, data=d, type="rr")
```

Or vice-versa

```{r   }
riskreg(y ~ a, nuisance=~z, propensity=~x+z, data=d, type="rr")
```

Whereas the MLE yields a biased estimate of the relative risk:

```{r   }
fit_mle <- with(d, riskreg_mle(y, a, x1=model.matrix(~1,d), x2=model.matrix(~z, d)))
estimate(fit_mle, 1)
```

To obtain an estimate of the risk-difference (here wrong model) we simply chance the `type` argument

```{r   }
riskreg(y ~ a, nuisance=~x+z, data=d, type="rd")
```


Interactions with the exposure can be examined with the `target` argument

```{r   }
riskreg(y ~ a, target=a~x, nuisance=~x+z, data=d, type="rr")
```



# Project organization

We use the `dev` branch for development and the `main` branch for stable
releases, which currently follow a frequency of about 4 weeks. All
releases follow [semantic versioning](https://semver.org/), are
[tagged](https://github.com/kkholst/targeted/tags) and notable
changes are reported in a
[changelog](https://github.com/kkholst/targeted/blob/main/NEWS.md).

# I Have a Question / I Want to Report a Bug

If you want to ask questions, require help or clarification, or report a
bug, we recommend to either contact a maintainer directly or the
following:

- Open an [Issue](https://github.com/kkholst/targeted/issues).
- Provide as much context as you can about what you’re running into.
- Provide project and platform versions, depending on what seems
  relevant.

We will then take care of the issue as soon as possible.
