---
# YAML header created by ox-ravel
title: Targeted inference (targeted)
author: Klaus K. Holst
---
<a href="https://travis-ci.org/kkholst/targeted"><img src="https://travis-ci.org/kkholst/targeted.svg?branch=master"></a>

<a href="https://codecov.io/github/kkholst/targeted?branch=master"><img src="https://codecov.io/github/kkholst/targeted/coverage.svg?branch=master"></a>

<a href="https://cran.r-project.org/package=targeted"><img src="http://www.r-pkg.org/badges/version/targeted"></a>

<a href="http://cranlogs.r-pkg.org/downloads/total/last-month/targeted"><img src="https://www.r-pkg.org/badges/version-last-release/targeted"></a>

<a href=""https://opensource.org/licenses/Apache-2.0"><img src="https://img.shields.io/badge/License-Apache%202.0-blue.svg"></a>


# Examples


## Targeted risk regression

```{r  eval=FALSE }
library(targeted)
library(magrittr)
```

Simulate some data:

```{r   }
m <- lvm() %>%
    regression(a ~ x+z) %>%
    regression(lp.target ~ 1) %>%
    regression(lp.nuisance ~ x + z) %>%
    distribution('a', binomial.lvm("logit")) %>%
    binomial.rr('y', 'a', 'lp.target', 'lp.nuisance')

par <- c('a'=-2, 'lp.target'=1, 'lp.nuisance'=-1, 'lp.nuisance~x'=2)
d <- lava::sim(m, n=1e4, seed=1, p=par) %>%
    subset(select=c('y', 'a','x','z'))

head(d)
```

```{r   }
fit <- riskreg(y ~ a, nuisance=~x+z, data=d, type="rr")
fit
```

Here the same design matrix is used for both the propensity model and
the nuisance parameter (odds-product) model

```{r   }
summary(fit)
```

Double-robustness illustrated by using a wrong propensity
model but a correct nuisance paramter (odds-product) model:

```{r   }
riskreg(y ~ a, nuisance=~x+z, propensity=~z, data=d, type="rr")
```

Or vice-versa

```{r   }
riskreg(y ~ a, nuisance=~z, propensity=~x+z, data=d, type="rr")
```

Whereas the MLE yields a biased estimate of the relative risk:

```{r   }
fit_mle <- with(d, riskreg_mle(y, a, x1=model.matrix(~1,d), x2=model.matrix(~z, d)))
estimate(fit_mle, 1)
```

To obtain an estimate of the risk-difference (here wrong model) we simply chance the `type` argument

```{r   }
riskreg(y ~ a, nuisance=~x+z, data=d, type="rd")
```


## Interactions

Interactions with the exposure can be examined with the `target` argument

```{r   }
riskreg(y ~ a, target=a~x, nuisance=~x+z, data=d, type="rr")
```
